---
title: '**Data Normalization: Index Sorting**'
jupyter: ir
---

## **Checking if data is linear or log**

### ERC_aim1_exp9_summary

```{r}

library(readxl)

df_9 <- read_excel("../data/ERC_aim1_exp9_summary.xlsx")
marker_cols_9 <- grep("_Median$", colnames(df_9), value = TRUE)

hist(df_9[[marker_cols_9[1]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_9[1])

hist(df_9[[marker_cols_9[2]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_9[2])

hist(df_9[[marker_cols_9[3]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_9[3])

hist(df_9[[marker_cols_9[4]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_9[4])
```

Log10 transformation

```{r}

log_data_1 <- log10(df_9[[marker_cols_9[1]]])
log_data_2 <- log10(df_9[[marker_cols_9[2]]])
log_data_3 <- log10(df_9[[marker_cols_9[3]]])

hist(log_data_1,
     breaks = 50,
     main = "Log-transformed data",
     xlab = paste0("log10(", marker_cols_9[1], ")"))

hist(log_data_2,
     breaks = 50,
     main = "Log-transformed data",
     xlab = paste0("log10(", marker_cols_9[2], ")"))

hist(log_data_3,
     breaks = 50,
     main = "Log-transformed data",
     xlab = paste0("log10(", marker_cols_9[3], ")"))
```

### ERC_aim1_exp15_summary

```{r}

df_15 <- read_excel("../data/ERC_aim1_exp15_summary.xlsx")
marker_cols_15 <- grep("_Median$", colnames(df_15), value = TRUE)
df_15[marker_cols_15] <- lapply(df_15[marker_cols_15], function(x) {
  as.numeric(gsub(",", ".", x))
})


hist(df_15[[marker_cols_15[1]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_15[1])

hist(df_15[[marker_cols_15[2]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_15[2])

hist(df_15[[marker_cols_15[3]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_15[3])

hist(df_15[[marker_cols_15[4]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_15[4])
```

Log10 transformation

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
log_data_1 <- log10(df_15[[marker_cols_15[1]]])
log_data_2 <- log10(df_15[[marker_cols_15[2]]])
log_data_3 <- log10(df_15[[marker_cols_15[3]]])

hist(log_data_1,
     breaks = 50,
     main = "Log-transformed data",
     xlab = paste0("log10(", marker_cols_15[1], ")"))

hist(log_data_2,
     breaks = 50,
     main = "Log-transformed data",
     xlab = paste0("log10(", marker_cols_15[2], ")"))

hist(log_data_3,
     breaks = 50,
     main = "Log-transformed data",
     xlab = paste0("log10(", marker_cols_15[3], ")"))
```

### ERC_aim_exp13_summary

```{r}
library(tidyverse)

#| colab: {base_uri: https://localhost:8080/, height: 371}
#| collapsed: true
library(readxl)
df_13 <- read_excel("../data/ERC_aim1_exp13_summary1.xlsx")
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
marker_cols_13 <- grep("_Median$", colnames(df_13), value = TRUE)
marker_cols_13

str(df_13[marker_cols_13[1]])

df_13[marker_cols_13] <- lapply(df_13[marker_cols_13], function(x) {
  as.numeric(gsub(",", ".", x))
})

hist(df_13[[marker_cols_13[1]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_13[1])

hist(df_13[[marker_cols_13[2]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_13[2])

hist(df_13[[marker_cols_13[3]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_13[3])

hist(df_13[[marker_cols_13[4]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_13[4])
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
log_data_1 <- log10(df_13[[marker_cols_13[1]]])
log_data_2 <- log10(df_13[[marker_cols_13[2]]])
log_data_3 <- log10(df_13[[marker_cols_13[3]]])

hist(log_data_1,
     breaks = 50,
     main = "Log-transformed data",
     xlab = paste0("log10(", marker_cols_13[1], ")"))

hist(log_data_2,
     breaks = 50,
     main = "Log-transformed data",
     xlab = paste0("log10(", marker_cols_13[2], ")"))

hist(log_data_3,
     breaks = 50,
     main = "Log-transformed data",
     xlab = paste0("log10(", marker_cols_13[3], ")"))
```

### Mateusz_summary

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 306}
#| collapsed: true
library(readxl)
df_mat <- read_excel("../data/Mateusz_summary.xlsx")
head(df_mat)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
marker_cols_mat <- grep("^Median", colnames(df_mat), value = TRUE)

str(df_mat[marker_cols_mat[1]])

df_mat[marker_cols_mat] <- lapply(df_mat[marker_cols_mat], function(x) {
  as.numeric(gsub(",", ".", x))
})

hist(df_mat[[marker_cols_mat[2]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_mat[2])

hist(df_mat[[marker_cols_mat[4]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_mat[4])

hist(df_mat[[marker_cols_mat[6]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_mat[6])

#all rows of these 3 columns have the same value=97776,47473
#we exclude them from further analysis as they do not give any biological information

marker_cols_mat <- marker_cols_mat[! marker_cols_mat %in% c('Median.1', 'Median.3', 'Median.5')]

hist(df_mat[[marker_cols_mat[1]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_mat[1])

hist(df_mat[[marker_cols_mat[2]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_mat[2])

hist(df_mat[[marker_cols_mat[3]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_mat[3])

hist(df_mat[[marker_cols_mat[4]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_mat[4])
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
log_data_1 <- log10(df_mat[[marker_cols_mat[1]]])
log_data_2 <- log10(df_mat[[marker_cols_mat[3]]])
log_data_3 <- log10(df_mat[[marker_cols_mat[5]]])

hist(log_data_1,
     breaks = 50,
     main = "Log-transformed data",
     xlab = paste0("log10(", marker_cols_mat[1], ")"))

hist(log_data_2,
     breaks = 50,
     main = "Log-transformed data",
     xlab = paste0("log10(", marker_cols_mat[3], ")"))

hist(log_data_3,
     breaks = 50,
     main = "Log-transformed data",
     xlab = paste0("log10(", marker_cols_mat[5], ")"))
```

### RSM

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 337}
library(readxl)
df_rsm <- read_excel("../data/RSM_summary.xlsx")
head(df_rsm)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
marker_cols_rsm <- grep("_Median$", colnames(df_rsm), value = TRUE)
marker_cols_rsm

str(df_rsm[marker_cols_rsm[1]])

df_rsm[marker_cols_rsm] <- lapply(df_rsm[marker_cols_rsm], function(x) {
  as.numeric(gsub(",", ".", x))
})

hist(df_rsm[[marker_cols_rsm[1]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_rsm[1])

hist(df_rsm[[marker_cols_rsm[2]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_rsm[2])

hist(df_rsm[[marker_cols_rsm[3]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_rsm[3])

hist(df_rsm[[marker_cols_rsm[4]]],
     breaks = 80,
     main = "Raw data",
     xlab = marker_cols_rsm[4])
```

log10

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
log_data_1 <- log10(df_rsm[[marker_cols_rsm[1]]])
log_data_2 <- log10(df_rsm[[marker_cols_rsm[2]]])
log_data_3 <- log10(df_rsm[[marker_cols_rsm[3]]])

hist(log_data_1,
     breaks = 50,
     main = "Log-transformed data",
     xlab = paste0("log10(", marker_cols_rsm[1], ")"))

hist(log_data_2,
     breaks = 50,
     main = "Log-transformed data",
     xlab = paste0("log10(", marker_cols_rsm[2], ")"))

hist(log_data_3,
     breaks = 50,
     main = "Log-transformed data",
     xlab = paste0("log10(", marker_cols_rsm[3], ")"))
```

###As we can see after logarithm transformation the distirbution is way better

## **Log transformation**

### Comparing log_10, ln, log1p and asinh

#### Comparing for marker from ERC_aim1_exp_9

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
#| collapsed: true
#log 10
df_log_9 <- df_9
df_log_9[marker_cols_9] <- log10(df_9[marker_cols_9])

hist(df_9[[marker_cols_9[8]]],
     breaks = 50,
     main = "before log - Raw data",
     xlab = marker_cols_9[8]
     )

hist(df_log_9[[marker_cols_9[8]]],
     breaks = 50,
     main = "after log10",
     xlab = paste0("log10(", marker_cols_9[8], ")")
     )

#ln
df_log_n <- df_9
df_log_n[marker_cols_9] <- log(df_9[marker_cols_9])


hist(df_log_n[[marker_cols_9[8]]],
     breaks = 50,
     main = "after log (natural)",
     xlab = paste0("log(", marker_cols_9[8], ")")
)

#log 1p
df_log_1p <- df_9
df_log_1p[marker_cols_9] <- log1p(df_9[marker_cols_9])


hist(df_log_1p[[marker_cols_9[8]]],
     breaks = 50,
     main = "after log 1p",
     xlab = paste0("log(", marker_cols_9[8], ")")
)

#asinh
df_log_asinh <- df_9
df_log_asinh[marker_cols_9] <- asinh(df_9[marker_cols_9] / 150)


hist(df_log_asinh[[marker_cols_9[8]]],
     breaks = 50,
     main = "after asinh",
     xlab = paste0("log(", marker_cols_9[8], ")")
)
```

####comparing for marker from file ERC_aim1_exp15

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
#| collapsed: true
#log_10
df_log_15 <- df_15
df_log_15[marker_cols_15] <- log10(df_15[marker_cols_15])

hist(df_15[[marker_cols_15[5]]],
     breaks = 50,
     main = "before log - Raw data",
     xlab = marker_cols_15[5]
     )

hist(df_log_15[[marker_cols_15[5]]],
     breaks = 50,
     main = "after log10",
     xlab = paste0("log10(", marker_cols_15[5], ")")
     )

#ln
df_log_n <- df_15
df_log_n[marker_cols_15] <- log(df_15[marker_cols_15])

hist(df_log_n[[marker_cols_15[5]]],
     breaks = 50,
     main = "after log (natural)",
     xlab = paste0("log(", marker_cols_15[5], ")")
)

#log 1p
df_log_1p <- df_15
df_log_1p[marker_cols_15] <- log1p(df_15[marker_cols_15])

hist(df_log_1p[[marker_cols_15[5]]],
     breaks = 50,
     main = "after log 1p",
     xlab = paste0("log(", marker_cols_15[5], ")")
)

#asinh
df_log_asinh <- df_15
df_log_asinh[marker_cols_15] <- asinh(df_15[marker_cols_15] / 150)

hist(df_log_asinh[[marker_cols_15[5]]],
     breaks = 50,
     main = "after asinh",
     xlab = paste0("log(", marker_cols_15[5], ")")
)
```

We decide on using log_10 in furter analysis as it seem to work best

### Log_10 transformation ERC_aim1_exp_9

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
#| collapsed: true
for (m in marker_cols_9) {
  hist((df_log_9[[m]]),
       breaks = 80,
       main = paste("After log transformation:", m),
       xlab = paste0("log(", marker_cols_9[m], ")"))
}
```

### Log_10 transformation ERC_aim1_exp_15

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
for (m in marker_cols_15) {
  hist((df_log_15[[m]]),
       breaks = 80,
       main = paste("After log transformation:", m),
       xlab = paste0("log(", marker_cols_15[m], ")"))
}
```

### Log_10 transformation ERC_aim1_exp_13

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
#| collapsed: true
df_log_13 <- df_13
df_log_13[marker_cols_13] <- log10(df_13[marker_cols_13])

for (m in marker_cols_13) {
  hist((df_log_13[[m]]),
       breaks = 80,
       main = paste("After log transformation:", m),
       xlab = paste0("log(", marker_cols_13[m], ")"))
}
```

### Log_10 transformation Mateusz

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
#| collapsed: true
df_log_mat <- df_mat
df_log_mat[marker_cols_mat] <- log10(df_mat[marker_cols_mat])

for (m in marker_cols_mat) {
  hist((df_log_mat[[m]]),
       breaks = 80,
       main = paste("After log transformation:", m),
       xlab = paste0("log(", m, ")"))
}
```

### Log_10 transformation rsm

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
df_log_rsm <- df_rsm
df_log_rsm[marker_cols_rsm] <- log10(df_rsm[marker_cols_rsm])

for (m in marker_cols_rsm) {
  hist((df_log_rsm[[m]]),
       breaks = 80,
       main = paste("After log transformation:", m),
       xlab = paste0("log(", m, ")"))
}
```

## **Z-score scalling**

### scalling ERC_aim1_exp_9

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 877}
#| collapsed: true
df_scaled_9 <- df_log_9
df_scaled_9[marker_cols_9] <- scale(df_log_9[marker_cols_9])
boxplot(df_scaled_9[marker_cols_9],
        main="log10 + z-score (scaled)",
        las=2)

hist(df_scaled_9[[marker_cols_9[8]]],
     breaks = 80,
     main = "after scalling",
     xlab = paste0("scale(", marker_cols_9[8], ")")
)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
#| collapsed: true
for (m in marker_cols_9) {
  hist(df_9[[m]],
       breaks = 80,
       main = paste("Before normalization:", m),
       xlab = m)

  hist((df_scaled_9[[m]]),
       breaks = 80,
       main = paste("After scalling:", m),
       xlab = paste0("log(", m, ")"))
}
```

### scalling ERC_aim1_exp_15

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 877}
#| collapsed: true
df_scaled_15 <- df_log_15
df_scaled_15[marker_cols_15] <- scale(df_log_15[marker_cols_15])
boxplot(df_scaled_15[marker_cols_15],
        main="log10 + z-score (scaled)",
        las=2)

hist(df_scaled_15[[marker_cols_15[8]]],
     breaks = 80,
     main = "after scalling",
     xlab = paste0("scale(", marker_cols_15[8], ")")
)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
#| collapsed: true
for (m in marker_cols_15) {
  hist(df_15[[m]],
       breaks = 80,
       main = paste("Before normalization:", m),
       xlab = m)

  hist((df_scaled_15[[m]]),
       breaks = 80,
       main = paste("After scalling:", m),
       xlab = paste0("log(", m, ")"))
}
```

### scalling ERC_aim1_exp_13

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 857}
df_scaled_13 <- df_log_13
df_scaled_13[marker_cols_13] <- scale(df_log_13[marker_cols_13])
boxplot(df_scaled_13[marker_cols_13],
        main="log10 + z-score (scaled)",
        las=2)

#here we choose numeric columns and then replace all NA for zero
num_cols <- sapply(df_scaled_13, is.numeric)
df_scaled_13[, num_cols] <- lapply(df_scaled_13[, num_cols], function(x) {
  x[is.na(x) | is.infinite(x)] <- 0
  x
})

hist(df_scaled_13[[marker_cols_13[8]]],
     breaks = 80,
     main = "after scalling",
     xlab = paste0("scale(", marker_cols_13[8], ")")
)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
for (m in marker_cols_13) {
  hist(df_13[[m]],
       breaks = 80,
       main = paste("Before normalization:", m),
       xlab = m)

  hist((df_scaled_13[[m]]),
       breaks = 80,
       main = paste("After scalling:", m),
       xlab = paste0("log(", m, ")"))
}
```

### scalling Mateusz

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 857}
df_scaled_mat <- df_log_mat
df_scaled_mat[marker_cols_mat] <- scale(df_log_mat[marker_cols_mat])
boxplot(df_scaled_mat[marker_cols_mat],
        main="log10 + z-score (scaled)",
        las=2)

#here we choose numeric columns and then replace all NA for zero
num_cols <- sapply(df_scaled_mat, is.numeric)
df_scaled_mat[, num_cols] <- lapply(df_scaled_mat[, num_cols], function(x) {
  x[is.na(x) | is.infinite(x)] <- 0
  x
})

hist(df_scaled_mat[[marker_cols_mat[8]]],
     breaks = 80,
     main = "after scalling",
     xlab = paste0("scale(", marker_cols_mat[8], ")")
)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
for (m in marker_cols_mat) {
  hist(df_mat[[m]],
       breaks = 80,
       main = paste("Before normalization:", m),
       xlab = m)

  hist((df_scaled_mat[[m]]),
       breaks = 80,
       main = paste("After scalling:", m),
       xlab = paste0("log(", m, ")"))
}
```

### scalling RSM

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 857}
df_scaled_rsm <- df_log_rsm
df_scaled_rsm[marker_cols_rsm] <- scale(df_log_rsm[marker_cols_rsm])
boxplot(df_scaled_rsm[marker_cols_rsm],
        main="log10 + z-score (scaled)",
        las=2)

#here we choose numeric columns and then replace all NA for zero
num_cols <- sapply(df_scaled_rsm, is.numeric)
df_scaled_rsm[, num_cols] <- lapply(df_scaled_rsm[, num_cols], function(x) {
  x[is.na(x) | is.infinite(x)] <- 0
  x
})

hist(df_scaled_rsm[[marker_cols_rsm[8]]],
     breaks = 80,
     main = "after scalling",
     xlab = paste0("scale(", marker_cols_rsm[8], ")")
)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
for (m in marker_cols_rsm) {
  hist(df_rsm[[m]],
       breaks = 80,
       main = paste("Before normalization:", m),
       xlab = m)

  hist((df_scaled_rsm[[m]]),
       breaks = 80,
       main = paste("After scalling:", m),
       xlab = paste0("log(", m, ")"))
}
```

## **PCA**

#### PCA exp9

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
#| collapsed: true
pca_9 <- prcomp(df_scaled_9[marker_cols_9])
plot(pca_9$x[,1], pca_9$x[,2])
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 34}
#| collapsed: true
pca_var_9 <- pca_9$sdev^2 # how much variation in the original data each principal component accounts for
pca_var_per_9 <- round(pca_var_9/sum(pca_var_9)*100,1)
pca_var_per_9
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
#| collapsed: true
barplot( pca_var_per_9, main = "scree plot", xlab = "principal component", ylab = "percent variation")
```

slight difference beetwen 2 and 3, maybe PC1 vs PC3 or PC2 vs PC3? delete outliers?

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
#| collapsed: true
library(ggplot2)

pca_9.data <- data.frame(Sample = df_9$Row.names,
  X = pca_9$x[,1],
  Y = pca_9$x[,2])

ggplot(data = pca_9.data, aes(x = X, y = Y, label = Sample))+
  geom_point(size = 1)+
  xlab(paste("PC1 -", pca_var_per_9[1], "%", sep = ""))+
  ylab(paste("PC2 -", pca_var_per_9[2], "%", sep = ""))+
  theme_bw()+
  ggtitle("PCA graph")
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 252}
loading_scores_9 <- pca_9$rotation[,1]
marker_scores_9 <- abs(loading_scores_9)
markers_score_ranked_9 <- sort(marker_scores_9, decreasing = TRUE)
top_5_markers_9 <- names(markers_score_ranked_9[1:5])

top5_loadings_9 <- data.frame(
  Marker = top_5_markers_9,
  Loading = pca_9$rotation[top_5_markers_9, 1] #show the scores and sign (+/-)
)

top5_loadings_9
```

#### Applying z-score colours on PCA exp9

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 675}
#| collapsed: true
library(dplyr)
library(tidyr)
library(ggplot2)

pca_9.data <- data.frame(
  Sample = rownames(df_9),
  X = pca_9$x[,1],
  Y = pca_9$x[,2]
)

plot_data <- pca_9.data %>%
  bind_cols(df_scaled_9[marker_cols_9]) %>%
  pivot_longer(
    cols = all_of(marker_cols_9),
    names_to = "Marker",
    values_to = "z_score"
  )

ggplot(plot_data, aes(x = X, y = Y, color = z_score)) +
  geom_point(size = 1) +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0
  ) +
  facet_wrap(~ Marker) +
  xlab(paste("PC1 -", pca_var_per_9[1], "%")) +
  ylab(paste("PC2 -", pca_var_per_9[2], "%")) +
  theme_bw() +
  ggtitle("PCA with z-score colours for all markers")
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
#| collapsed: true
for (marker in marker_cols_9) {
  pca_9.data$z_var9 <- df_scaled_9[[marker]]
  p <- ggplot(data = pca_9.data,
            aes(
              x = X,
              y = Y,
              label = Sample,
              color = z_var9)
          ) +
            scale_color_gradient2(
              low = "blue",
              mid = "white",
              high = "red",
              midpoint = 0,
              name = "z-score"
          ) +
            geom_point(size = 1)+
            xlab(paste("PC1 -", pca_var_per_9[1], "%", sep = ""))+
            ylab(paste("PC2 -", pca_var_per_9[2], "%", sep = ""))+
            theme_bw()+
            ggtitle(marker)
  print(p)
}
```

#### PCA exp15

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
#| collapsed: true
pca_15 <- prcomp(df_scaled_15[marker_cols_15])
plot(pca_15$x[,1], pca_15$x[,2])
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 34}
pca_var_15 <- pca_15$sdev^2 # how much variation in the original data each principal component accounts for
pca_var_per_15 <- round(pca_var_15 / sum(pca_var_15)*100,1)
pca_var_per_15
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
#| collapsed: true
barplot( pca_var_per_15, main = "scree plot", xlab = "principal component", ylab = "percent variation")
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
#| collapsed: true
library(ggplot2)

pca_15.data <- data.frame(Sample = 1:nrow(df_15),
  X = pca_15$x[,1],
  Y = pca_15$x[,2])

ggplot(data = pca_15.data, aes(x = X, y = Y, label = Sample))+
  geom_point(size = 1)+
  xlab(paste("PC1 -", pca_var_per_15[1], "%", sep = ""))+
  ylab(paste("PC2 -", pca_var_per_15[2], "%", sep = ""))+
  theme_bw()+
  ggtitle("PCA graph")
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 252}
#| collapsed: true
loading_scores_15 <- pca_15$rotation[,1]
marker_scores_15 <- abs(loading_scores_15)
markers_score_ranked_15 <- sort(marker_scores_15, decreasing = TRUE)
top_5_markers_15 <- names(markers_score_ranked_15[1:5])

top5_loadings_15 <- data.frame(
  Marker = top_5_markers_15,
  Loading = pca_15$rotation[top_5_markers_15, 1] #show the scores and sign (+/-)
)

top5_loadings_15
```

#### Applying z-score colours on PCA exp15

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
#| collapsed: true
library(dplyr)
library(tidyr)
library(ggplot2)

pca_15.data <- data.frame(
  Sample = rownames(df_15),
  X = pca_15$x[,1],
  Y = pca_15$x[,2]
)

plot_data <- pca_15.data %>%
  bind_cols(df_scaled_15[marker_cols_15]) %>%
  pivot_longer(
    cols = all_of(marker_cols_15),
    names_to = "Marker",
    values_to = "z_score"
  )

ggplot(plot_data, aes(x = X, y = Y, color = z_score)) +
  geom_point(size = 1) +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0
  ) +
  facet_wrap(~ Marker) +
  xlab(paste("PC1 -", pca_var_per_15[1], "%")) +
  ylab(paste("PC2 -", pca_var_per_15[2], "%")) +
  theme_bw() +
  ggtitle("PCA with z-score colours for all markers")
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
#| collapsed: true
for (marker in marker_cols_15) {
  pca_15.data$z_var15 <- df_scaled_15[[marker]]
  p <- ggplot(data = pca_15.data,
            aes(
              x = X,
              y = Y,
              label = Sample,
              color = z_var15)
          ) +
            scale_color_gradient2(
              low = "blue",
              mid = "white",
              high = "red",
              midpoint = 0,
              name = "z-score"
          ) +
            geom_point(size = 1)+
            xlab(paste("PC1 -", pca_var_per_15[1], "%", sep = ""))+
            ylab(paste("PC2 -", pca_var_per_15[2], "%", sep = ""))+
            theme_bw()+
            ggtitle(marker)
  print(p)
}
```

#### PCA exp13

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}

pca_13 <- prcomp(df_scaled_13[marker_cols_13])
plot(pca_13$x[,1], pca_13$x[,2])
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 34}
pca_var_13 <- pca_13$sdev^2 # how much variation in the original data each principal component accounts for
pca_var_per_13 <- round(pca_var_13/sum(pca_var_13)*100,1)
pca_var_per_13
```

```{r}
sum(pca_var_per_13[1:8])
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
barplot( pca_var_per_13, main = "scree plot", xlab = "principal component", ylab = "percent variation")
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
library(ggplot2)

pca_13.data <- data.frame(Sample = df_13$Row.names,
  X = pca_13$x[,1],
  Y = pca_13$x[,2])

ggplot(data = pca_13.data, aes(x = X, y = Y, label = Sample))+
  geom_point(size = 1)+
  xlab(paste("PC1 -", pca_var_per_13[1], "%", sep = ""))+
  ylab(paste("PC2 -", pca_var_per_13[2], "%", sep = ""))+
  theme_bw()+
  ggtitle("PCA graph")
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 252}
loading_scores_13 <- pca_13$rotation[,1]
marker_scores_13 <- abs(loading_scores_13)
markers_score_ranked_13 <- sort(marker_scores_13, decreasing = TRUE)
top_5_markers_13 <- names(markers_score_ranked_13[1:5])

top5_loadings_13 <- data.frame(
  Marker = top_5_markers_13,
  Loading = pca_13$rotation[top_5_markers_13, 1] #show the scores and sign (+/-)
)

top5_loadings_13
```

#### Applying z-score colours on PCA exp13

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
#| collapsed: true
library(dplyr)
library(tidyr)
library(ggplot2)

pca_13.data <- data.frame(
  Sample = rownames(df_13),
  X = pca_13$x[,1],
  Y = pca_13$x[,2]
)

plot_data <- pca_13.data %>%
  bind_cols(df_scaled_13[marker_cols_13]) %>%
  pivot_longer(
    cols = all_of(marker_cols_13),
    names_to = "Marker",
    values_to = "z_score"
  )

ggplot(plot_data, aes(x = X, y = Y, color = z_score)) +
  geom_point(size = 1) +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0
  ) +
  facet_wrap(~ Marker) +
  xlab(paste("PC1 -", pca_var_per_13[1], "%")) +
  ylab(paste("PC2 -", pca_var_per_13[2], "%")) +
  theme_bw() +
  ggtitle("PCA with z-score colours for all markers")
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
#| collapsed: true
for (marker in marker_cols_13) {
  pca_13.data$z_var13 <- df_scaled_13[[marker]]
  p <- ggplot(data = pca_13.data,
            aes(
              x = X,
              y = Y,
              label = Sample,
              color = z_var13)
          ) +
            scale_color_gradient2(
              low = "blue",
              mid = "white",
              high = "red",
              midpoint = 0,
              name = "z-score"
          ) +
            geom_point(size = 1)+
            xlab(paste("PC1 -", pca_var_per_13[1], "%", sep = ""))+
            ylab(paste("PC2 -", pca_var_per_13[2], "%", sep = ""))+
            theme_bw()+
            ggtitle(marker)
  print(p)
}
```

#### PCA Mateusz

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
pca_mat <- prcomp(df_scaled_mat[marker_cols_mat])
plot(pca_mat$x[,1], pca_mat$x[,2])
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 34}
pca_var_mat <- pca_mat$sdev^2 # how much variation in the original data each principal component accounts for
pca_var_per_mat <- round(pca_var_mat/sum(pca_var_mat)*100,1)
pca_var_per_mat
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
barplot(pca_var_per_mat, main = "scree plot", xlab = "principal component", ylab = "percent variation")
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
library(ggplot2)

pca_mat.data <- data.frame(Sample = rownames(df_mat),
  X = pca_mat$x[,1],
  Y = pca_mat$x[,2])

ggplot(data = pca_mat.data, aes(x = X, y = Y, label = Sample))+
  geom_point(size = 1)+
  xlab(paste("PC1 -", pca_var_per_mat[1], "%", sep = ""))+
  ylab(paste("PC2 -", pca_var_per_mat[2], "%", sep = ""))+
  theme_bw()+
  ggtitle("PCA graph")
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 255}
loading_scores_mat <- pca_mat$rotation[,1]
marker_scores_mat <- abs(loading_scores_mat)
markers_score_ranked_mat <- sort(marker_scores_mat, decreasing = TRUE)
top_5_markers_mat <- names(markers_score_ranked_mat[1:5])

top5_loadings_mat <- data.frame(
  Marker = top_5_markers_mat,
  Loading = pca_mat$rotation[top_5_markers_mat, 1] #show the scores and sign (+/-)
)

top5_loadings_mat
```

#### Applying z-score colours on PCA Mateusz

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
#| collapsed: true
library(dplyr)
library(tidyr)
library(ggplot2)

pca_mat.data <- data.frame(
  Sample = rownames(df_mat),
  X = pca_mat$x[,1],
  Y = pca_mat$x[,2]
)

plot_data <- pca_mat.data %>%
  bind_cols(df_scaled_mat[marker_cols_mat]) %>%
  pivot_longer(
    cols = all_of(marker_cols_mat),
    names_to = "Marker",
    values_to = "z_score"
  )

ggplot(plot_data, aes(x = X, y = Y, color = z_score)) +
  geom_point(size = 1) +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0
  ) +
  facet_wrap(~ Marker) +
  xlab(paste("PC1 -", pca_var_per_mat[1], "%")) +
  ylab(paste("PC2 -", pca_var_per_mat[2], "%")) +
  theme_bw() +
  ggtitle("PCA with z-score colours for all markers")
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
#| collapsed: true
for (marker in marker_cols_mat) {
  pca_mat.data$z_var_mat <- df_scaled_mat[[marker]]
  p <- ggplot(data = pca_mat.data,
            aes(
              x = X,
              y = Y,
              label = Sample,
              color = z_var_mat)
          ) +
            scale_color_gradient2(
              low = "blue",
              mid = "white",
              high = "red",
              midpoint = 0,
              name = "z-score"
          ) +
            geom_point(size = 1)+
            xlab(paste("PC1 -", pca_var_per_mat[1], "%", sep = ""))+
            ylab(paste("PC2 -", pca_var_per_mat[2], "%", sep = ""))+
            theme_bw()+
            ggtitle(marker)
  print(p)
}
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
pca_15.data$z_var15 <- df_scaled_15[[ marker_cols_15[8]]]

ggplot(data = pca_15.data,
        aes(
          x = X,
          y = Y,
          label = Sample,
          color = z_var15)
      ) +
        scale_color_gradient2(
          low = "blue",
          mid = "white",
          high = "red",
          midpoint = 0,
          name = "Z-score"
      ) +
        geom_point(size = 1)+
        xlab(paste("PC1 -", pca_var_per_15[1], "%", sep = ""))+
        ylab(paste("PC2 -", pca_var_per_15[2], "%", sep = ""))+
        theme_bw()+
        ggtitle("PCA graph with z-score colours")
```

### PCA RSM

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
pca_rsm <- prcomp(df_scaled_rsm[marker_cols_rsm])
plot(pca_rsm$x[,1], pca_rsm$x[,2])
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 34}
pca_var_rsm <- pca_rsm$sdev^2 # how much variation in the original data each principal component accounts for
pca_var_per_rsm <- round(pca_var_rsm/sum(pca_var_rsm)*100,1)
pca_var_per_rsm
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
barplot( pca_var_per_rsm, main = "scree plot", xlab = "principal component", ylab = "percent variation")
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
library(ggplot2)

pca_rsm.data <- data.frame(Sample = rownames(df_rsm),
  X = pca_rsm$x[,1],
  Y = pca_rsm$x[,2])

ggplot(data = pca_rsm.data, aes(x = X, y = Y, label = Sample))+
  geom_point(size = 1)+
  xlab(paste("PC1 -", pca_var_per_rsm[1], "%", sep = ""))+
  ylab(paste("PC2 -", pca_var_per_rsm[2], "%", sep = ""))+
  theme_bw()+
  ggtitle("PCA graph")
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 252}
loading_scores_rsm <- pca_rsm$rotation[,1]
marker_scores_rsm <- abs(loading_scores_rsm)
markers_score_ranked_rsm <- sort(marker_scores_rsm, decreasing = TRUE)
top_5_markers_rsm <- names(markers_score_ranked_rsm[1:5])

top5_loadings_rsm <- data.frame(
  Marker = top_5_markers_rsm,
  Loading = pca_rsm$rotation[top_5_markers_rsm, 1] #show the scores and sign (+/-)
)

top5_loadings_rsm
```

### Applying z-score colours on PCA rsm

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 675}
library(dplyr)
library(tidyr)
library(ggplot2)

pca_rsm.data <- data.frame(
  Sample = rownames(df_rsm),
  X = pca_rsm$x[,1],
  Y = pca_rsm$x[,2]
)

plot_data <- pca_rsm.data %>%
  bind_cols(df_scaled_rsm[marker_cols_rsm]) %>%
  pivot_longer(
    cols = all_of(marker_cols_rsm),
    names_to = "Marker",
    values_to = "z_score"
  )

ggplot(plot_data, aes(x = X, y = Y, color = z_score)) +
  geom_point(size = 1) +
  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0
  ) +
  facet_wrap(~ Marker) +
  xlab(paste("PC1 -", pca_var_per_rsm[1], "%")) +
  ylab(paste("PC2 -", pca_var_per_rsm[2], "%")) +
  theme_bw() +
  ggtitle("PCA with z-score colours for all markers")
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
for (marker in marker_cols_rsm) {
  pca_rsm.data$z_varrsm <- df_scaled_rsm[[marker]]
  p <- ggplot(data = pca_rsm.data,
            aes(
              x = X,
              y = Y,
              label = Sample,
              color = z_varrsm)
          ) +
            scale_color_gradient2(
              low = "blue",
              mid = "white",
              high = "red",
              midpoint = 0,
              name = "z-score"
          ) +
            geom_point(size = 1)+
            xlab(paste("PC1 -", pca_var_per_rsm[1], "%", sep = ""))+
            ylab(paste("PC2 -", pca_var_per_rsm[2], "%", sep = ""))+
            theme_bw()+
            ggtitle(marker)
  print(p)
}
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
pca_rsm.data$z_varrsm <- df_scaled_rsm[[ marker_cols_rsm[8]]]

ggplot(data = pca_rsm.data,
        aes(
          x = X,
          y = Y,
          label = Sample,
          color = z_varrsm)
      ) +
        scale_color_gradient2(
          low = "blue",
          mid = "white",
          high = "red",
          midpoint = 0,
          name = "Z-score"
      ) +
        geom_point(size = 1)+
        xlab(paste("PC1 -", pca_var_per_rsm[1], "%", sep = ""))+
        ylab(paste("PC2 -", pca_var_per_rsm[2], "%", sep = ""))+
        theme_bw()+
        ggtitle("PCA graph with z-score colours")
```

## **3D PCA**

```{r}
#| colab: {base_uri: https://localhost:8080/}
install.packages("scatterplot3d")
library(scatterplot3d)
```

### 3D PCA exp9

```{r}
pca_9.data <- data.frame(
  Sample = rownames(df_9),
  PC1 = pca_9$x[,1],
  PC2 = pca_9$x[,2],
  PC3 = pca_9$x[,3]
)

pca_var_9 <- pca_9$sdev^2
pca_var_per_9 <- round(pca_var_9 / sum(pca_var_9) * 100, 1)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
for (marker in marker_cols_9) {

  pca_9.data$z_var9 <- df_scaled_9[[marker]]


  cols <- colorRampPalette(c("blue", "white", "red"))(100)
  col_index <- as.numeric(cut(pca_9.data$z_var9, breaks = 100))

  scatterplot3d(
    x = pca_9.data$PC1,
    y = pca_9.data$PC2,
    z = pca_9.data$PC3,
    pch = 16,
    color = cols[col_index],
    xlab = paste0("PC1 - ", pca_var_per_9[1], "%"),
    ylab = paste0("PC2 - ", pca_var_per_9[2], "%"),
    zlab = paste0("PC3 - ", pca_var_per_9[3], "%"),
    main = marker
  )
}
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 34}
unique(df_9$GROUP)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
pca_9.data <- data.frame(
  Sample = rownames(df_9),
  PC1 = pca_9$x[,1],
  PC2 = pca_9$x[,2],
  PC3 = pca_9$x[,3],
  Group = df_9$GROUP
)

pca_var_9 <- pca_9$sdev^2
pca_var_per_9 <- round(pca_var_9 / sum(pca_var_9) * 100, 1)


group_colors <- c(
  "NaCl_woGCSF" = "blue",
  "NaCl_GCSF" = "green",
  "GCSF_GCSF" = "red",
  "GCSF_woGCSF" = "orange"
)


scatterplot3d(
  x = pca_9.data$PC1,
  y = pca_9.data$PC2,
  z = pca_9.data$PC3,
  pch = 16,
  color = group_colors[pca_9.data$Group],
  xlab = paste0("PC1 - ", pca_var_per_9[1], "%"),
  ylab = paste0("PC2 - ", pca_var_per_9[2], "%"),
  zlab = paste0("PC3 - ", pca_var_per_9[3], "%"),
  main = "PCA 3D colored by Group"
)

legend("topright", legend = names(group_colors), col = group_colors, pch = 16)
```

### 3D PCA exp15

```{r}
pca_15.data <- data.frame(
  Sample = rownames(df_15),
  PC1 = pca_15$x[,1],
  PC2 = pca_15$x[,2],
  PC3 = pca_15$x[,3]
)

pca_var_15 <- pca_15$sdev^2
pca_var_per_15 <- round(pca_var_15 / sum(pca_var_15) * 100, 1)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
for (marker in marker_cols_15) {

  pca_15.data$z_var15 <- df_scaled_15[[marker]]


  cols <- colorRampPalette(c("blue", "white", "red"))(100)
  col_index <- as.numeric(cut(pca_15.data$z_var15, breaks = 100))

  scatterplot3d(
    x = pca_15.data$PC1,
    y = pca_15.data$PC2,
    z = pca_15.data$PC3,
    pch = 16,
    color = cols[col_index],
    xlab = paste0("PC1 - ", pca_var_per_15[1], "%"),
    ylab = paste0("PC2 - ", pca_var_per_15[2], "%"),
    zlab = paste0("PC3 - ", pca_var_per_15[3], "%"),
    main = marker
  )
}
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 102}
colnames(df_15)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 34}
unique(df_15$treatment)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
pca_15.data <- data.frame(
  Sample = rownames(df_15),
  PC1 = pca_15$x[,1],
  PC2 = pca_15$x[,2],
  PC3 = pca_15$x[,3],
  Group = df_15$treatment
)

pca_var_15 <- pca_15$sdev^2
pca_var_per_15 <- round(pca_var_15 / sum(pca_var_15) * 100, 1)


group_colors <- c(
  "NaCl" = "blue",
  "PHZ" = "green",
  "GCSF" = "red"
)


scatterplot3d(
  x = pca_15.data$PC1,
  y = pca_15.data$PC2,
  z = pca_15.data$PC3,
  pch = 16,
  color = group_colors[pca_15.data$Group],
  xlab = paste0("PC1 - ", pca_var_per_15[1], "%"),
  ylab = paste0("PC2 - ", pca_var_per_15[2], "%"),
  zlab = paste0("PC3 - ", pca_var_per_15[3], "%"),
  main = "PCA 3D colored by Group"
)

legend("topright", legend = names(group_colors), col = group_colors, pch = 16)
```

### 3D PCA exp13

```{r}
pca_13.data <- data.frame(
  Sample = rownames(df_13),
  PC1 = pca_13$x[,1],
  PC2 = pca_13$x[,2],
  PC3 = pca_13$x[,3]
)

pca_var_13 <- pca_13$sdev^2
pca_var_per_13 <- round(pca_var_13 / sum(pca_var_13) * 100, 1)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
for (marker in marker_cols_13) {

  pca_13.data$z_var13 <- df_scaled_13[[marker]]


  cols <- colorRampPalette(c("blue", "white", "red"))(100)
  col_index <- as.numeric(cut(pca_13.data$z_var13, breaks = 100))

  scatterplot3d(
    x = pca_13.data$PC1,
    y = pca_13.data$PC2,
    z = pca_13.data$PC3,
    pch = 16,
    color = cols[col_index],
    xlab = paste0("PC1 - ", pca_var_per_13[1], "%"),
    ylab = paste0("PC2 - ", pca_var_per_13[2], "%"),
    zlab = paste0("PC3 - ", pca_var_per_13[3], "%"),
    main = marker
  )
}
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 272}
colnames(df_13)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 34}
unique(df_13$GROUP)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
pca_13.data <- data.frame(
  Sample = rownames(df_13),
  PC1 = pca_13$x[,1],
  PC2 = pca_13$x[,2],
  PC3 = pca_13$x[,3],
  Group = df_13$GROUP
)

pca_var_13 <- pca_13$sdev^2
pca_var_per_13 <- round(pca_var_13 / sum(pca_var_13) * 100, 1)


group_colors <- c(
  "N_N_woGCSF" = "blue",
  "N_N_GCSF" = "green",
  "G_N_woGCSF" = "red",
  "G_N_GCSF" = "orange",
  "G_G_GCSF" = "purple",
  "G_G_woGCSF" = "yellow",
  "N_G_woGCSF" = "black",
  "N_G_GCSF" = "brown",
  "NA" = "pink"
)


scatterplot3d(
  x = pca_13.data$PC1,
  y = pca_13.data$PC2,
  z = pca_13.data$PC3,
  pch = 16,
  color = group_colors[pca_13.data$Group],
  xlab = paste0("PC1 - ", pca_var_per_13[1], "%"),
  ylab = paste0("PC2 - ", pca_var_per_13[2], "%"),
  zlab = paste0("PC3 - ", pca_var_per_13[3], "%"),
  main = "PCA 3D colored by Group"
)

legend("topright", legend = names(group_colors), col = group_colors, pch = 16)
```

### 3D PCA Mateusz

```{r}
pca_mat.data <- data.frame(
  Sample = rownames(df_mat),
  PC1 = pca_mat$x[,1],
  PC2 = pca_mat$x[,2],
  PC3 = pca_mat$x[,3]
)

pca_var_mat <- pca_mat$sdev^2
pca_var_per_mat <- round(pca_var_mat / sum(pca_var_mat) * 100, 1)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
for (marker in marker_cols_mat) {

  pca_mat.data$z_var_mat<- df_scaled_mat[[marker]]


  cols <- colorRampPalette(c("blue", "white", "red"))(100)
  col_index <- as.numeric(cut(pca_mat.data$z_var_mat, breaks = 100))

  scatterplot3d(
    x = pca_mat.data$PC1,
    y = pca_mat.data$PC2,
    z = pca_mat.data$PC3,
    pch = 16,
    color = cols[col_index],
    xlab = paste0("PC1 - ", pca_var_per_mat[1], "%"),
    ylab = paste0("PC2 - ", pca_var_per_mat[2], "%"),
    zlab = paste0("PC3 - ", pca_var_per_mat[3], "%"),
    main = marker
  )
}
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 173}
colnames(df_mat)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 34}
unique(df_mat$Group.x)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 437}
pca_mat.data <- data.frame(
  Sample = rownames(df_mat),
  PC1 = pca_mat$x[,1],
  PC2 = pca_mat$x[,2],
  PC3 = pca_mat$x[,3],
  Group = df_mat$Group.x
)

pca_var_mat <- pca_mat$sdev^2
pca_var_per_mat <- round(pca_var_mat / sum(pca_var_mat) * 100, 1)


group_colors <- c(
  "Y_Neo_neg" = "blue",
  "Y_Neo_pos" = "green",
  "O_Neo_neg" = "red",
  "O_Neo_pos" = "orange"
)


scatterplot3d(
  x = pca_mat.data$PC1,
  y = pca_mat.data$PC2,
  z = pca_mat.data$PC3,
  pch = 16,
  color = group_colors[pca_mat.data$Group],
  xlab = paste0("PC1 - ", pca_var_per_mat[1], "%"),
  ylab = paste0("PC2 - ", pca_var_per_mat[2], "%"),
  zlab = paste0("PC3 - ", pca_var_per_mat[3], "%"),
  main = "PCA 3D colored by Group"
)

legend("topright", legend = names(group_colors), col = group_colors, pch = 16)
```

### 3D PCA rsm

```{r}
pca_rsm.data <- data.frame(
  Sample = rownames(df_rsm),
  PC1 = pca_rsm$x[,1],
  PC2 = pca_rsm$x[,2],
  PC3 = pca_rsm$x[,3]
)

pca_var_rsm <- pca_rsm$sdev^2
pca_var_per_rsm <- round(pca_var_rsm / sum(pca_var_rsm) * 100, 1)
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 1000}
for (marker in marker_cols_rsm) {

  pca_rsm.data$z_varrsm <- df_scaled_rsm[[marker]]


  cols <- colorRampPalette(c("blue", "white", "red"))(100)
  col_index <- as.numeric(cut(pca_rsm.data$z_varrsm, breaks = 100))

  scatterplot3d(
    x = pca_rsm.data$PC1,
    y = pca_rsm.data$PC2,
    z = pca_rsm.data$PC3,
    pch = 16,
    color = cols[col_index],
    xlab = paste0("PC1 - ", pca_var_per_rsm[1], "%"),
    ylab = paste0("PC2 - ", pca_var_per_rsm[2], "%"),
    zlab = paste0("PC3 - ", pca_var_per_rsm[3], "%"),
    main = marker
  )
}
```

```{r}
#| colab: {base_uri: https://localhost:8080/, height: 255}
colnames(df_rsm)
```

## Louvain/Leiden clustering

```{r}
pca_13.data <- data.frame(
  Sample = df_13$Row.names,
  X      = pca_13$x[, 1],
  Y      = pca_13$x[, 2]
)

ggplot(pca_13.data, aes(x = X, y = Y, label = Sample)) +
  geom_point(size = 1) +
  xlab(paste0("PC1 - ", pca_var_per_13[1], "%")) +
  ylab(paste0("PC2 - ", pca_var_per_13[2], "%")) +
  theme_bw() +
  ggtitle("PCA")
```

```{r}
#install.packages("leidenAlg")
library(igraph)
library(FNN)
library(leidenAlg)
```

```{r}
#which(cumsum(pca_var_per_13) >= 80)[1]
```

```{r}

find_elbow <- function(var_per) {
  n  <- length(var_per)
  x1 <- 1;  y1 <- var_per[1]
  x2 <- n;  y2 <- var_per[n]
  
  distances <- sapply(1:n, function(i) {
    abs((y2-y1)*i - (x2-x1)*var_per[i] + x2*y1 - y2*x1) /
      sqrt((y2-y1)^2 + (x2-x1)^2)
  })
  which.max(distances)
}

elbow <- find_elbow(pca_var_per_13)
cat("Łokieć w PC:", elbow, "\n")

# ── Scree plot z zaznaczonym łokciem ──────────────────────────────────────────

plot(pca_var_per_13,
     type = "b", pch = 19,
     xlab = "PC", ylab = "% wariancji",
     main = "Scree plot")
points(elbow, pca_var_per_13[elbow], col = "red", pch = 19, cex = 2)
abline(v = elbow, col = "red", lty = 2)
```

```{r}
sum(pca_var_per_13[1:elbow])
length(pca_var_per_13)
```

Mniej niż 80% wariancji, pozbywamy się szumu ale również wielu infromacji biologicznych ALE najwyższe modularity właśnie przy PC 4

```{r}
n_pcs      <- elbow # 7 
pca_coords <- pca_13$x[, 1:n_pcs]

k          <- 15
knn_result <- get.knn(pca_coords, k = k)

edges <- do.call(rbind, lapply(1:nrow(knn_result$nn.index), function(i) {
  cbind(i, knn_result$nn.index[i, ])
}))

g <- graph_from_edgelist(edges, directed = FALSE)
g <- simplify(g)

# Louvain clustering ────────────────────────────────────────────────────────
set.seed(123)
louvain_result <- cluster_louvain(g)
df_scaled_13$cluster_louvain <- as.factor(membership(louvain_result))
cat("Liczba klastrów (Louvain):", length(unique(membership(louvain_result))), "\n")

# Leiden clustering ───────────────────────────────────────────
set.seed(123)
leiden_result <- leiden.community(g, resolution = 0.5)

df_scaled_13$cluster_leiden <- as.factor(membership(leiden_result))
cat("Liczba klastrów (Leiden):", length(unique(membership(leiden_result))), "\n")

# Leiden clustering 1 ───────────────────────────────────────────
set.seed(123)
leiden_result_1 <- leiden.community(g, resolution = 1)

df_scaled_13$cluster_leiden_1 <- as.factor(membership(leiden_result_1))
cat("Liczba klastrów (Leiden):", length(unique(membership(leiden_result_1))), "\n")

# Wizualizacja klastrów na PCA ──────────────────────────────────────────────


pca_13.data$cluster_louvain <- df_scaled_13$cluster_louvain
pca_13.data$cluster_leiden  <- df_scaled_13$cluster_leiden
pca_13.data$cluster_leiden_1  <- df_scaled_13$cluster_leiden_1

ggplot(pca_13.data, aes(x = X, y = Y, color = cluster_louvain)) +
  geom_point(size = 1, alpha = 0.7) +
  theme_bw() +
  ggtitle("exp 13: Louvain clustering na PCA") +
  labs(color = "Klaster") +
  xlab(paste0("PC1 - ", pca_var_per_13[1], "%")) +
  ylab(paste0("PC2 - ", pca_var_per_13[2], "%"))

ggplot(pca_13.data, aes(x = X, y = Y, color = cluster_leiden)) +
  geom_point(size = 1, alpha = 0.7) +
  theme_bw() +
  ggtitle("exp 13: Leiden clustering na PCA (resolution 0.5)") +
  labs(color = "Klaster") +
  xlab(paste0("PC1 - ", pca_var_per_13[1], "%")) +
  ylab(paste0("PC2 - ", pca_var_per_13[2], "%"))

ggplot(pca_13.data, aes(x = X, y = Y, color = cluster_leiden_1)) +
  geom_point(size = 1, alpha = 0.7) +
  theme_bw() +
  ggtitle("exp 13: Leiden clustering na PCA (resolution 1) ") +
  labs(color = "Klaster") +
  xlab(paste0("PC1 - ", pca_var_per_13[1], "%")) +
  ylab(paste0("PC2 - ", pca_var_per_13[2], "%"))

```

Jakie resolution przy leiden?

Resolution w algorytmie Leiden kontroluje stopień szczegółowości klastrowania grafów/społeczności. Wyższe wartości (np. \> 1) prowadzą do powstania większej liczby mniejszych klastrów, podczas gdy niższe wartości (np. \< 1) generują mniej, ale większych społeczności.

Stabilność:

```{r}
table(df_scaled_13$cluster_louvain)

table(df_scaled_13$cluster_leiden)
table(df_scaled_13$cluster_leiden_1)
```

```{r}
modularity(louvain_result)

modularity(g, membership(leiden_result))
modularity(g, membership(leiden_result_1))
```

leiden przy resolution 1 daje największą modularność

### Louvain/Leiden exp 9

```{r}

```


## UMPAPs

### UMAP exp9
```{r}
cumvar_9 <- cumsum(pca_var_per_9)
cumvar_9

# Ile PC potrzeba do 80% wariancji
n_pc_9 <- which(cumvar_9 >= 80)[1]
cat("PC potrzebne do wyjaśnienia 80% wariancji:", n_pc_9, "\n")
cat("Skumulowana wariancja dla", n_pc_9, "PC:", cumvar_9[n_pc_9], "%\n")
```

```{r}
library(uwot)


pca_data_9 <- pca_9$x[, 1:8, drop = FALSE]

# UMAP
umap_9 <- uwot::umap(pca_data_9)

# do ramki danych
umap_df_9 <- data.frame(
  UMAP1 = umap_9[,1],
  UMAP2 = umap_9[,2]
)

head(umap_df_9)
```
```{r}
ggplot(umap_df_9, aes(x = UMAP1, y = UMAP2)) +
  geom_point(size = 2) +
  theme_minimal()
```

### UMAP exp15
```{r}
cumvar_15 <- cumsum(pca_var_per_15)
cumvar_15

# Ile PC potrzeba do 80% wariancji
n_pc_15 <- which(cumvar_15 >= 80)[1]
cat("PC potrzebne do wyjaśnienia 80% wariancji:", n_pc_15, "\n")
cat("Skumulowana wariancja dla", n_pc_15, "PC:", cumvar_15[n_pc_15], "%\n")
```

```{r}
library(uwot)


pca_data_15 <- pca_15$x[, 1:7, drop = FALSE]

# UMAP
umap_15 <- uwot::umap(pca_data_15)

# do ramki danych
umap_df_15 <- data.frame(
  UMAP1 = umap_15[,1],
  UMAP2 = umap_15[,2]
)
```

```{r}
ggplot(umap_df_15, aes(x = UMAP1, y = UMAP2)) +
  geom_point(size = 2) +
  theme_minimal()
```

### UMAP exp13

```{r}
cumvar_13 <- cumsum(pca_var_per_13)
cumvar_13

# Ile PC potrzeba
n_pc_13 <- which(cumvar_13 >= 80)[1]
cat("PC potrzebne do wyjaśnienia 80% wariancji:", n_pc_13, "\n")
cat("Skumulowana wariancja dla", n_pc_13, "PC:", cumvar_13[n_pc_13], "%\n")
```

```{r}
library(uwot)


pca_data_13 <- pca_13$x[, 1:7, drop = FALSE]

# UMAP
umap_13 <- uwot::umap(pca_data_13)

# do ramki danych
umap_df_13 <- data.frame(
  UMAP1 = umap_13[,1],
  UMAP2 = umap_13[,2]
)
```

```{r}
ggplot(umap_df_13, aes(x = UMAP1, y = UMAP2)) +
  geom_point(size = 2) +
  theme_minimal()
```

### UMAP exp_Mateusz

```{r}
cumvar_mat <- cumsum(pca_var_per_mat)
cumvar_mat

# Ile PC potrzeba do 80% wariancji
n_pc_mat <- which(cumvar_mat >= 80)[1]
cat("PC potrzebne do wyjaśnienia 80% wariancji:", n_pc_mat, "\n")
cat("Skumulowana wariancja dla", n_pc_mat, "PC:", cumvar_mat[n_pc_mat], "%\n")
```

```{r}
library(uwot)


pca_data_mat <- pca_mat$x[, 1:7, drop = FALSE]

# UMAP
umap_mat <- uwot::umap(pca_data_mat)

# do ramki danych
umap_df_mat <- data.frame(
  UMAP1 = umap_mat[,1],
  UMAP2 = umap_mat[,2]
)
```

```{r}
ggplot(umap_df_mat, aes(x = UMAP1, y = UMAP2)) +
  geom_point(size = 2) +
  theme_minimal()
```


### UMAP exp_rsm

```{r}
cumvar_rsm <- cumsum(pca_var_per_rsm)
cumvar_rsm

# Ile PC potrzeba do 80% wariancji
n_pc_rsm <- which(cumvar_rsm >= 80)[1]
cat("PC potrzebne do wyjaśnienia 80% wariancji:", n_pc_rsm, "\n")
cat("Skumulowana wariancja dla", n_pc_rsm, "PC:", cumvar_rsm[n_pc_rsm], "%\n")
```

```{r}
library(uwot)


pca_data_rsm <- pca_rsm$x[, 1:5, drop = FALSE]

# UMAP
umap_rsm <- uwot::umap(pca_data_rsm)

# do ramki danych
umap_df_rsm <- data.frame(
  UMAP1 = umap_rsm[,1],
  UMAP2 = umap_rsm[,2]
)
```

```{r}
ggplot(umap_df_rsm, aes(x = UMAP1, y = UMAP2)) +
  geom_point(size = 2) +
  theme_minimal()
```
